---
title: "04 – Negative & Positive Controls QC"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 4
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: false
---

# Overview

This notebook assesses the quality of **negative controls** (extraction blanks,
PCR no-template controls) and **positive controls** (mock communities), and
removes contaminants identified from negative controls before downstream analysis.

**Negative controls:** Detect and remove ASVs introduced by reagent or
laboratory contamination using the `decontam` package.

**Positive controls:** Compare observed composition to the expected mock
community composition to assess sequencing accuracy and DADA2 performance.

---

# 1. User Parameters

```{r}
#| label: params
# ── Paths ─────────────────────────────────────────────────────────────────────
dada2_dir     <- here::here("data", "dada2")
metadata_path <- here::here("data", "metadata", "metadata.csv")
phyloseq_dir  <- here::here("data", "phyloseq")

# ── Metadata column names ─────────────────────────────────────────────────────
sample_id_col   <- "SampleID"    # Column with sample identifiers
sample_type_col <- "SampleType"  # Column indicating "Sample", "NegControl", "PosControl"
dna_conc_col    <- "DNAConc_ng"  # (Optional) DNA concentration for frequency method

# ── decontam settings ─────────────────────────────────────────────────────────
decontam_method    <- "prevalence"  # "prevalence", "frequency", or "combined"
decontam_threshold <- 0.1           # Probability threshold for calling contaminants

# ── Positive control: expected composition ────────────────────────────────────
# Provide expected genus-level composition of your mock community.
# Leave NULL to skip mock community evaluation.
# Example: ZymoBIOMICS D6305 mock (8 bacterial, 2 fungal species – bacteria only shown)
mock_expected <- data.frame(
  Genus        = c("Listeria", "Pseudomonas", "Bacillus", "Lactobacillus",
                   "Escherichia", "Salmonella", "Enterococcus", "Staphylococcus"),
  Expected_pct = c(12, 12, 12, 12, 12, 12, 12, 12),
  stringsAsFactors = FALSE
)
# Set NULL if no positive control in your experiment:
# mock_expected <- NULL
```

---

# 2. Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(phyloseq)
  library(decontam)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(patchwork)
  library(here)
})

dir.create(phyloseq_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# 3. Load Data

```{r}
#| label: load-dada2
seqtab_nochim <- readRDS(file.path(dada2_dir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(dada2_dir, "taxa.rds"))
metadata      <- read.csv(metadata_path, stringsAsFactors = FALSE)
rownames(metadata) <- metadata[[sample_id_col]]

cat("Samples in seqtab:", nrow(seqtab_nochim), "\n")
cat("Samples in metadata:", nrow(metadata), "\n")
cat("ASVs:", ncol(seqtab_nochim), "\n")
```

```{r}
#| label: sample-type-check
if (!sample_type_col %in% colnames(metadata)) {
  stop("Column '", sample_type_col, "' not found in metadata.\n",
       "Please add a column indicating 'Sample', 'NegControl', or 'PosControl'.")
}

cat("\nSample type counts:\n")
print(table(metadata[[sample_type_col]]))
```

```{r}
#| label: build-ps-raw
# Build a preliminary phyloseq object for decontam
# (no tree yet – that comes in notebook 05)
ps_raw <- phyloseq(
  otu_table(seqtab_nochim, taxa_are_rows = FALSE),
  tax_table(taxa),
  sample_data(metadata)
)

# Rename ASVs for readability
dna <- Biostrings::DNAStringSet(taxa_names(ps_raw))
names(dna) <- paste0("ASV", seq_along(dna))
taxa_names(ps_raw) <- names(dna)

ps_raw
```

---

# 4. Negative Controls – Contamination Detection

## 4.1 Identify Negative Controls

```{r}
#| label: id-neg-controls
sample_data(ps_raw)$is_neg <- sample_data(ps_raw)[[sample_type_col]] == "NegControl"

n_neg <- sum(sample_data(ps_raw)$is_neg)
n_pos <- sum(sample_data(ps_raw)[[sample_type_col]] == "PosControl")
n_smp <- sum(sample_data(ps_raw)[[sample_type_col]] == "Sample")

cat(sprintf("  True samples:      %d\n", n_smp))
cat(sprintf("  Negative controls: %d\n", n_neg))
cat(sprintf("  Positive controls: %d\n", n_pos))

if (n_neg == 0) {
  warning("No negative controls found. Skipping decontam analysis.")
}
```

## 4.2 Read Count in Negative Controls

```{r}
#| label: neg-control-reads
#| fig-width: 8
#| fig-height: 4
read_counts <- data.frame(
  Sample = sample_names(ps_raw),
  Reads  = sample_sums(ps_raw),
  Type   = sample_data(ps_raw)[[sample_type_col]],
  stringsAsFactors = FALSE
)

ggplot(read_counts, aes(x = reorder(Sample, Reads), y = Reads, fill = Type)) +
  geom_col() +
  scale_fill_manual(values = c(Sample = "steelblue", NegControl = "tomato",
                                PosControl = "goldenrod")) +
  scale_y_log10(labels = scales::comma) +
  coord_flip() +
  labs(title = "Read counts per sample",
       x = NULL, y = "Reads (log10)", fill = "Type") +
  theme_bw()
```

## 4.3 Run decontam

```{r}
#| label: run-decontam
if (n_neg > 0) {
  if (decontam_method %in% c("frequency", "combined")) {
    if (!dna_conc_col %in% colnames(sample_data(ps_raw))) {
      warning("DNA concentration column '", dna_conc_col, "' not found.\n",
              "Falling back to prevalence method.")
      decontam_method <- "prevalence"
    } else {
      sample_data(ps_raw)$quant_reading <- sample_data(ps_raw)[[dna_conc_col]]
    }
  }

  contam_result <- isContaminant(
    ps_raw,
    method    = decontam_method,
    neg       = "is_neg",
    threshold = decontam_threshold,
    detailed  = TRUE
  )

  n_contam <- sum(contam_result$contaminant, na.rm = TRUE)
  cat(sprintf("\nContaminants identified: %d / %d ASVs (%.1f%%)\n",
              n_contam, nrow(contam_result),
              100 * n_contam / nrow(contam_result)))
}
```

## 4.4 Inspect Contaminant Scores

```{r}
#| label: decontam-scores-plot
#| fig-width: 7
#| fig-height: 4
if (n_neg > 0 && exists("contam_result")) {
  ggplot(contam_result, aes(x = p, fill = contaminant)) +
    geom_histogram(bins = 30, colour = "white") +
    scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
                      labels = c("Not contaminant", "Contaminant")) +
    geom_vline(xintercept = decontam_threshold, linetype = "dashed") +
    labs(title = "decontam – contaminant probability distribution",
         x = "p (probability of being a contaminant)",
         y = "Number of ASVs",
         fill = NULL) +
    theme_bw()
}
```

## 4.5 Inspect Contaminant Taxa

```{r}
#| label: contam-taxa-table
if (n_neg > 0 && exists("contam_result")) {
  contam_asvs <- rownames(contam_result)[contam_result$contaminant & !is.na(contam_result$contaminant)]

  if (length(contam_asvs) > 0) {
    contam_tax <- as.data.frame(tax_table(ps_raw)[contam_asvs, ])
    contam_tax$ASV <- contam_asvs
    contam_tax$p_contam <- contam_result[contam_asvs, "p"]
    knitr::kable(contam_tax[order(contam_tax$p_contam), ],
                 caption = "Putative contaminant ASVs",
                 row.names = FALSE)
  } else {
    cat("No contaminants detected at threshold p <", decontam_threshold, "\n")
  }
}
```

## 4.6 Prevalence Pattern: Contaminants vs True Taxa

A true contaminant should appear more prevalently in negative controls
(low DNA input) relative to true samples.

```{r}
#| label: prevalence-plot
#| fig-width: 7
#| fig-height: 5
if (n_neg > 0 && exists("contam_result") && n_contam > 0) {
  ps_neg <- prune_samples(sample_data(ps_raw)$is_neg, ps_raw)
  ps_smp <- prune_samples(!sample_data(ps_raw)$is_neg, ps_raw)

  prev_neg <- apply(otu_table(ps_neg), 2, function(x) mean(x > 0))
  prev_smp <- apply(otu_table(ps_smp), 2, function(x) mean(x > 0))

  prev_df <- data.frame(
    ASV         = names(prev_neg),
    Prev_neg    = prev_neg,
    Prev_sample = prev_smp,
    Contaminant = contam_result[names(prev_neg), "contaminant"],
    stringsAsFactors = FALSE
  )

  ggplot(prev_df, aes(x = Prev_neg, y = Prev_sample,
                       colour = Contaminant)) +
    geom_point(alpha = 0.6) +
    scale_colour_manual(values = c("FALSE" = "steelblue", "TRUE" = "tomato")) +
    labs(title = "ASV prevalence: negative controls vs samples",
         x = "Prevalence in negative controls",
         y = "Prevalence in true samples",
         colour = "Contaminant") +
    theme_bw()
}
```

---

# 5. Remove Contaminants

```{r}
#| label: remove-contaminants
if (n_neg > 0 && exists("contam_result") && n_contam > 0) {
  ps_clean <- prune_taxa(!contam_result$contaminant | is.na(contam_result$contaminant),
                          ps_raw)
  cat(sprintf("ASVs before removal: %d\n", ntaxa(ps_raw)))
  cat(sprintf("ASVs after  removal: %d\n", ntaxa(ps_clean)))
} else {
  ps_clean <- ps_raw
  message("No contaminants removed.")
}

# Remove negative controls from downstream analysis
ps_nocontrols <- subset_samples(ps_clean,
                                 sample_data(ps_clean)[[sample_type_col]] != "NegControl")
cat(sprintf("\nSamples after removing negative controls: %d\n", nsamples(ps_nocontrols)))
```

---

# 6. Positive Controls – Mock Community Assessment

## 6.1 Extract Mock Samples

```{r}
#| label: extract-mock
if (!is.null(mock_expected) &&
    any(sample_data(ps_clean)[[sample_type_col]] == "PosControl")) {

  ps_mock <- subset_samples(ps_clean,
                              sample_data(ps_clean)[[sample_type_col]] == "PosControl")
  cat("Mock community samples:", nsamples(ps_mock), "\n")
  cat("ASVs detected in mock:", ntaxa(prune_taxa(taxa_sums(ps_mock) > 0, ps_mock)), "\n")
} else {
  message("No positive controls found or expected composition not provided. Skipping mock assessment.")
  ps_mock <- NULL
}
```

## 6.2 Genus-level Composition vs Expected

```{r}
#| label: mock-composition
#| fig-width: 8
#| fig-height: 5
if (!is.null(ps_mock)) {
  # Aggregate to genus level
  ps_mock_genus <- tax_glom(ps_mock, "Genus")
  mock_rel <- transform_sample_counts(ps_mock_genus, function(x) 100 * x / sum(x))

  mock_obs_df <- psmelt(mock_rel) %>%
    group_by(Genus) %>%
    summarise(Observed_pct = mean(Abundance), .groups = "drop") %>%
    filter(Observed_pct >= 1)  # Only show genera with >=1%

  # Merge with expected
  mock_compare <- merge(mock_obs_df, mock_expected, by = "Genus", all = TRUE)
  mock_compare[is.na(mock_compare)] <- 0

  mock_long <- tidyr::pivot_longer(
    mock_compare,
    cols      = c("Observed_pct", "Expected_pct"),
    names_to  = "Source",
    values_to = "Percent"
  )

  ggplot(mock_long, aes(x = reorder(Genus, Percent),
                         y = Percent,
                         fill = Source)) +
    geom_col(position = "dodge") +
    coord_flip() +
    scale_fill_manual(values = c(Expected_pct = "grey60", Observed_pct = "steelblue")) +
    labs(title = "Positive control: observed vs expected composition",
         x = NULL, y = "Relative abundance (%)", fill = NULL) +
    theme_bw()
}
```

## 6.3 Sequence Accuracy: Error Rate Estimate

```{r}
#| label: mock-error-rate
if (!is.null(ps_mock)) {
  mock_asvs_rel <- transform_sample_counts(ps_mock, function(x) x / sum(x))
  mock_asv_tab  <- otu_table(mock_asvs_rel)

  total_mock_asvs <- sum(colSums(as.data.frame(otu_table(ps_mock))) > 0)
  expected_genera <- mock_expected$Genus
  mock_tax        <- as.data.frame(tax_table(ps_mock))

  true_asvs  <- sum(mock_tax$Genus %in% expected_genera &
                      taxa_sums(ps_mock) > 0)
  error_asvs <- total_mock_asvs - true_asvs

  cat(sprintf("Total ASVs in mock:            %d\n", total_mock_asvs))
  cat(sprintf("ASVs matching expected genera: %d\n", true_asvs))
  cat(sprintf("Spurious ASVs (potential errors): %d (%.1f%%)\n",
              error_asvs, 100 * error_asvs / total_mock_asvs))
}
```

---

# 7. Save Cleaned Phyloseq Object

```{r}
#| label: save-ps-clean
ps_for_downstream <- ps_nocontrols

# Also remove positive controls
ps_for_downstream <- subset_samples(
  ps_for_downstream,
  sample_data(ps_for_downstream)[[sample_type_col]] == "Sample"
)

# Drop ASVs with zero counts after sample removal
ps_for_downstream <- prune_taxa(taxa_sums(ps_for_downstream) > 0, ps_for_downstream)

cat(sprintf("\nFinal dataset after QC:\n"))
cat(sprintf("  Samples: %d\n", nsamples(ps_for_downstream)))
cat(sprintf("  ASVs:    %d\n", ntaxa(ps_for_downstream)))

saveRDS(ps_for_downstream, file.path(phyloseq_dir, "ps_after_controls_qc.rds"))
message("Cleaned phyloseq object saved to: ",
        file.path(phyloseq_dir, "ps_after_controls_qc.rds"))
```

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
