---
title: "05 – phyloseq Assembly & Filtering"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 4
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: false
---

# Overview

This notebook builds a clean, analysis-ready **phyloseq** object from the
decontaminated DADA2 outputs. Steps:

1. Assemble the phyloseq object (OTU table, taxonomy, sample data)
2. (Optional) Build a phylogenetic tree with DECIPHER + phangorn
3. Prevalence and count-based filtering
4. Exploratory taxonomic overview
5. Save the final object

---

# 1. User Parameters

```{r}
#| label: params
# ── Paths ─────────────────────────────────────────────────────────────────────
dada2_dir        <- here::here("data", "dada2")
metadata_path    <- here::here("data", "metadata", "metadata.csv")
phyloseq_dir     <- here::here("data", "phyloseq")
controls_ps_path <- file.path(phyloseq_dir, "ps_after_controls_qc.rds")

# ── Metadata ──────────────────────────────────────────────────────────────────
sample_id_col <- "SampleID"  # Column with unique sample identifiers

# ── Phylogenetic tree construction ────────────────────────────────────────────
build_tree <- TRUE   # Set FALSE to skip (can be slow for large datasets)
n_cores    <- 4L

# ── Filtering criteria ────────────────────────────────────────────────────────
# Remove ASVs not present in at least `min_prevalence` samples
min_prevalence <- 2   # minimum number of samples an ASV must appear in

# Remove ASVs with fewer than `min_total_count` reads across all samples
min_total_count <- 5

# ── Taxonomy filters ──────────────────────────────────────────────────────────
# Remove these taxa (contaminants, off-target amplification)
remove_kingdoms <- c("Eukaryota", "NA")   # NA = unclassified kingdom
remove_phyla    <- c()                    # e.g. c("Chloroplast")

# SILVA-specific: remove chloroplast / mitochondria
remove_class  <- c("Chloroplast")
remove_family <- c("Mitochondria")
```

---

# 2. Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(phyloseq)
  library(DECIPHER)
  library(phangorn)
  library(ggplot2)
  library(dplyr)
  library(patchwork)
  library(here)
})

dir.create(phyloseq_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# 3. Load or Build phyloseq Object

## From controls QC output (recommended)

```{r}
#| label: load-from-controls-qc
if (file.exists(controls_ps_path)) {
  ps <- readRDS(controls_ps_path)
  message("Loaded phyloseq from controls QC: ", controls_ps_path)
  ps
} else {
  message("Controls QC phyloseq not found. Building from DADA2 outputs directly.")
  build_from_dada2 <- TRUE
}
```

## From DADA2 outputs (fallback if notebook 04 was skipped)

```{r}
#| label: build-from-dada2
#| eval: !expr !exists("ps")
seqtab_nochim <- readRDS(file.path(dada2_dir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(dada2_dir, "taxa.rds"))
metadata      <- read.csv(metadata_path, stringsAsFactors = FALSE)
rownames(metadata) <- metadata[[sample_id_col]]

ps <- phyloseq(
  otu_table(seqtab_nochim, taxa_are_rows = FALSE),
  tax_table(taxa),
  sample_data(metadata)
)

# Rename to ASV1, ASV2, ...
taxa_names(ps) <- paste0("ASV", seq_len(ntaxa(ps)))
ps
```

---

# 4. Phylogenetic Tree (Optional)

A phylogenetic tree enables UniFrac distance metrics and Faith's PD. We use
**DECIPHER** to perform a multiple sequence alignment and **phangorn** to
construct a maximum-likelihood tree (GTR + G model).

```{r}
#| label: build-tree
#| eval: !expr !build_tree
message("Building phylogenetic tree (this may take several minutes)...")

asv_seqs <- colnames(otu_table(ps))
if (taxa_are_rows(ps)) asv_seqs <- rownames(otu_table(ps))

# Use stored ASV names
names(asv_seqs) <- taxa_names(ps)

dna_seqs <- Biostrings::DNAStringSet(asv_seqs)

# Multiple sequence alignment
message("  Aligning sequences with DECIPHER...")
alignment <- DECIPHER::AlignSeqs(dna_seqs, anchor = NA, processors = n_cores)

# Convert to phyDat object for phangorn
phang_align <- phangorn::phyDat(as(alignment, "matrix"), type = "DNA")

# Initial NJ tree
message("  Building initial NJ tree...")
dm      <- phangorn::dist.ml(phang_align)
tree_nj <- phangorn::NJ(dm)

# Fit GTR + G model for ML tree
message("  Fitting GTR+G maximum-likelihood tree...")
fit_init     <- phangorn::pml(tree_nj, data = phang_align)
fit_gtrgamma <- phangorn::optim.pml(
  fit_init,
  model         = "GTR",
  optInv        = TRUE,
  optGamma      = TRUE,
  rearrangement = "stochastic",
  control       = phangorn::pml.control(trace = 0)
)

phy_tree <- phangorn::midpoint(fit_gtrgamma$tree)
message("  Tree constructed: ", phangorn::Nnode(phy_tree), " nodes")

# Add tree to phyloseq
ps <- merge_phyloseq(ps, phy_tree)
message("Phylogenetic tree added to phyloseq object.")
```

```{r}
#| label: tree-plot
#| eval: !expr !build_tree
#| fig-width: 6
#| fig-height: 8
plot_tree(ps, ladderize = "left", justify = "left",
          size = "Abundance", color = "Phylum") +
  ggtitle("Phylogenetic tree coloured by Phylum") +
  theme(legend.position = "right")
```

---

# 5. Taxonomic Filtering

Remove non-target sequences (chloroplasts, mitochondria, unclassified kingdoms).

```{r}
#| label: tax-filter
ps_tax <- ps

# Remove specified kingdoms
if (length(remove_kingdoms) > 0) {
  ps_tax <- subset_taxa(ps_tax,
    !is.na(Kingdom) & !(Kingdom %in% remove_kingdoms))
}

# Remove specific phyla
if (length(remove_phyla) > 0) {
  ps_tax <- subset_taxa(ps_tax, !(Phylum %in% remove_phyla))
}

# Remove chloroplasts (Class level)
if (length(remove_class) > 0) {
  ps_tax <- subset_taxa(ps_tax, !(Class %in% remove_class))
}

# Remove mitochondria (Family level)
if (length(remove_family) > 0) {
  ps_tax <- subset_taxa(ps_tax, !(Family %in% remove_family))
}

cat(sprintf("ASVs before taxonomic filtering: %d\n", ntaxa(ps)))
cat(sprintf("ASVs after  taxonomic filtering: %d\n", ntaxa(ps_tax)))
```

---

# 6. Prevalence and Count Filtering

```{r}
#| label: prevalence-filter
prev_df <- data.frame(
  ASV        = taxa_names(ps_tax),
  Prevalence = apply(otu_table(ps_tax), ifelse(taxa_are_rows(ps_tax), 1, 2),
                      function(x) sum(x > 0)),
  TotalCount = taxa_sums(ps_tax),
  stringsAsFactors = FALSE
)

ggplot(prev_df, aes(x = TotalCount, y = Prevalence)) +
  geom_point(alpha = 0.4, colour = "steelblue") +
  geom_vline(xintercept = min_total_count, linetype = "dashed", colour = "red") +
  geom_hline(yintercept = min_prevalence,  linetype = "dashed", colour = "orange") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "ASV prevalence vs total read count",
       subtitle = sprintf("Thresholds: prevalence ≥ %d, count ≥ %d",
                          min_prevalence, min_total_count),
       x = "Total reads (log10)", y = "Prevalence (# samples)") +
  theme_bw()
```

```{r}
#| label: apply-filter
keep_asvs <- prev_df$ASV[prev_df$Prevalence  >= min_prevalence &
                            prev_df$TotalCount >= min_total_count]
ps_filt <- prune_taxa(keep_asvs, ps_tax)

cat(sprintf("ASVs before prevalence/count filter: %d\n", ntaxa(ps_tax)))
cat(sprintf("ASVs after  prevalence/count filter: %d\n", ntaxa(ps_filt)))
cat(sprintf("Reads retained: %.1f%%\n",
            100 * sum(otu_table(ps_filt)) / sum(otu_table(ps_tax))))
```

---

# 7. Exploratory Taxonomic Overview

```{r}
#| label: phylum-composition
#| fig-width: 9
#| fig-height: 5
ps_rel    <- transform_sample_counts(ps_filt, function(x) x / sum(x))
ps_phylum <- tax_glom(ps_rel, "Phylum")

plot_bar(ps_phylum, fill = "Phylum") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Phylum-level relative abundance",
       y = "Relative abundance", x = "Sample") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")
```

```{r}
#| label: top-genera
#| fig-width: 9
#| fig-height: 5
ps_genus   <- tax_glom(ps_rel, "Genus")
top_genera <- names(sort(taxa_sums(ps_genus), decreasing = TRUE))[1:20]

ps_top <- prune_taxa(top_genera, ps_genus)

plot_bar(ps_top, fill = "Genus") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Top 20 genera by mean relative abundance",
       y = "Relative abundance", x = "Sample") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 8. Sequencing Depth Summary

```{r}
#| label: depth-summary
#| fig-width: 7
#| fig-height: 4
depth_df <- data.frame(
  Sample = sample_names(ps_filt),
  Reads  = sample_sums(ps_filt),
  stringsAsFactors = FALSE
)

ggplot(depth_df, aes(x = Reads)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = median(depth_df$Reads),
             linetype = "dashed", colour = "red") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Sequencing depth distribution",
       subtitle = sprintf("Median: %s reads", format(median(depth_df$Reads), big.mark=",")),
       x = "Reads per sample", y = "Count") +
  theme_bw()
```

---

# 9. Save Final phyloseq Object

```{r}
#| label: save-final
saveRDS(ps_filt, file.path(phyloseq_dir, "ps.rds"))

# Also save filtered ASV sequences as FASTA
actual_seqs <- colnames(otu_table(ps_filt))
names(actual_seqs) <- taxa_names(ps_filt)
Biostrings::writeXStringSet(
  Biostrings::DNAStringSet(actual_seqs),
  filepath = file.path(phyloseq_dir, "asvs_filtered.fasta")
)

cat("\nFinal phyloseq object:\n")
print(ps_filt)
message("\nSaved to: ", file.path(phyloseq_dir, "ps.rds"))
```

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
