---
title: "06 – Diversity Analysis"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 4
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: false
  fig-width: 8
  fig-height: 5
---

# Overview

This notebook computes alpha and beta diversity. A key feature is the use of
**bias-corrected** richness and diversity estimators:

| Estimator | Package | Bias correction |
|-----------|---------|----------------|
| Observed richness | base | None (underestimates true richness) |
| breakaway S | **breakaway** | Ratio estimator; robust to library depth |
| Shannon H′ | **DivNet** | Network-aware; accounts for correlation structure |
| Simpson (1–D) | **DivNet** | Network-aware |
| Faith's PD | phyloseq + picante | None; uses phylogenetic tree |

For beta diversity we use Bray-Curtis, weighted/unweighted UniFrac, and
Aitchison distance (CLR-transformed), with PERMANOVA and PERMDISP tests.

---

# 1. User Parameters

```{r}
#| label: params
# ── Paths ─────────────────────────────────────────────────────────────────────
phyloseq_dir <- here::here("data", "phyloseq")
ps_path      <- file.path(phyloseq_dir, "ps.rds")
results_dir  <- here::here("data", "diversity")

# ── Study design ──────────────────────────────────────────────────────────────
group_var      <- "Group"   # Metadata column for group comparisons
# (Optional) Covariates for DivNet / betta regression
covariate_vars <- c()       # e.g. c("Age", "Sex")

# ── Rarefaction ───────────────────────────────────────────────────────────────
# For standard (non-bias-corrected) metrics only.
# Set to NULL to use minimum library depth.
rarefy_depth <- NULL   # e.g. 5000; or NULL

# ── DivNet settings ───────────────────────────────────────────────────────────
# DivNet fits a network model; B controls bootstrap samples (default 5 is fast)
divnet_tuning <- "fast"  # "fast" (B=5) or "careful" (B=10)

# ── Statistical test for alpha diversity ──────────────────────────────────────
alpha_stat_test <- "betta"  # "betta" (breakaway), "wilcox", or "kruskal"
alpha_p_adjust  <- "BH"     # Method for multiple comparison correction

# ── Beta diversity ────────────────────────────────────────────────────────────
beta_distances <- c("bray", "wunifrac", "unifrac", "aitchison")
n_permutations <- 999   # PERMANOVA permutations
```

---

# 2. Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(phyloseq)
  library(breakaway)
  library(DivNet)
  library(vegan)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(patchwork)
  library(here)
})

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# 3. Load Data

```{r}
#| label: load-ps
ps <- readRDS(ps_path)
ps
```

```{r}
#| label: check-group-var
if (!group_var %in% colnames(sample_data(ps))) {
  stop("Group variable '", group_var, "' not found in sample metadata.\n",
       "Available columns: ", paste(colnames(sample_data(ps)), collapse = ", "))
}

groups <- sample_data(ps)[[group_var]]
cat("Groups:", paste(unique(groups), collapse = ", "), "\n")
cat("Group sizes:\n"); print(table(groups))
```

---

# 4. Alpha Diversity

## 4.1 Standard Metrics (Rarefied Data)

Standard alpha diversity metrics are computed on a rarefied (subsampled) table
to equalise library depth. Note that rarefaction discards data and is
controversial; bias-corrected estimates (below) are preferred.

```{r}
#| label: rarefy
set.seed(42)

if (is.null(rarefy_depth)) {
  rarefy_depth <- min(sample_sums(ps))
  message("Rarefying to minimum depth: ", format(rarefy_depth, big.mark=","))
}

ps_rare <- rarefy_even_depth(ps, sample.size = rarefy_depth,
                               rngseed = 42, verbose = FALSE)
cat(sprintf("Samples retained after rarefaction: %d\n", nsamples(ps_rare)))
```

```{r}
#| label: standard-alpha
#| fig-width: 10
#| fig-height: 5
alpha_std <- estimate_richness(ps_rare,
  measures = c("Observed", "Shannon", "Simpson", "Chao1", "ACE"))
alpha_std$Sample <- rownames(alpha_std)
alpha_std$Group  <- sample_data(ps_rare)[[group_var]]

alpha_long <- pivot_longer(alpha_std,
  cols      = c("Observed", "Shannon", "Simpson", "Chao1", "ACE"),
  names_to  = "Metric",
  values_to = "Value"
)

ggplot(alpha_long, aes_string(x = "Group", y = "Value", fill = "Group")) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.5) +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Standard alpha diversity metrics (rarefied)",
       x = NULL, y = "Diversity") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

## 4.2 Faith's Phylogenetic Diversity

Requires a phylogenetic tree in the phyloseq object.

```{r}
#| label: faiths-pd
if (!is.null(phy_tree(ps_rare, errorIfNULL = FALSE))) {
  # picante-style PD: sum of branch lengths
  otu_mat <- as.matrix(otu_table(ps_rare))
  if (!taxa_are_rows(ps_rare)) otu_mat <- t(otu_mat)

  pd_df <- picante::pd(t(otu_mat), phy_tree(ps_rare))
  pd_df$Group  <- sample_data(ps_rare)[[group_var]]
  pd_df$Sample <- rownames(pd_df)

  p_pd <- ggplot(pd_df, aes_string(x = "Group", y = "PD", fill = "Group")) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.6) +
    scale_fill_brewer(palette = "Set2") +
    labs(title = "Faith's Phylogenetic Diversity", x = NULL, y = "PD") +
    theme_bw() +
    theme(legend.position = "none")
  print(p_pd)
} else {
  message("No phylogenetic tree found – skipping Faith's PD.")
}
```

## 4.3 Bias-corrected Richness with breakaway

`breakaway` (Willis & Bunge, 2015) estimates true richness using frequency
ratios, providing estimates that are **not** sensitive to sequencing depth.

```{r}
#| label: breakaway-richness
message("Running breakaway richness estimation (this may take a few minutes)...")

# Run on the full (non-rarefied) table
ba_estimates <- breakaway::breakaway(ps)
ba_summary   <- breakaway::summary(ba_estimates)
ba_summary$Group  <- sample_data(ps)[[group_var]]
ba_summary$Sample <- sample_names(ps)

p_ba <- ggplot(ba_summary, aes_string(x = "Group", y = "estimate",
                                        colour = "Group")) +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  geom_point(position = position_jitter(0.15), size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.1, alpha = 0.4) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "breakaway: bias-corrected richness estimates",
       subtitle = "Error bars = 95% confidence interval",
       x = NULL, y = "Estimated richness") +
  theme_bw() +
  theme(legend.position = "none")

print(p_ba)
```

## 4.4 breakaway betta: Group Comparison

`betta()` (Willis *et al.*, 2017) regresses breakaway estimates on covariates,
properly accounting for the uncertainty in each richness estimate.

```{r}
#| label: betta-test
covariate_formula_str <- group_var
if (length(covariate_vars) > 0) {
  covariate_formula_str <- paste(c(group_var, covariate_vars), collapse = " + ")
}

meta_betta <- as.data.frame(sample_data(ps))

betta_result <- breakaway::betta(
  chats = ba_summary$estimate,
  ses   = ba_summary$std_error,
  X     = model.matrix(as.formula(paste("~", covariate_formula_str)),
                        data = meta_betta)
)

cat("\n── betta regression: richness ~ ", covariate_formula_str, " ──\n")
print(betta_result$table)
```

## 4.5 Bias-corrected Diversity with DivNet

`DivNet` (Willis & Martin, 2020) estimates Shannon entropy and other diversity
metrics while accounting for correlation between taxa using a network model.
It provides variance estimates and enables proper hypothesis testing.

```{r}
#| label: divnet-shannon
message("Running DivNet (Shannon + Simpson)...")
message("  This fits a network model and may take several minutes.")

# DivNet requires the non-rarefied phyloseq object
dv <- DivNet::divnet(
  W      = ps,
  X      = group_var,   # grouping variable; set to NULL for no covariate
  tuning = divnet_tuning
)

# Extract Shannon estimates
shannon_dv <- dv$shannon %>%
  DivNet::summary() %>%
  as.data.frame()

shannon_dv$Group  <- sample_data(ps)[[group_var]]
shannon_dv$Sample <- sample_names(ps)

p_shannon <- ggplot(shannon_dv,
                     aes_string(x = "Group", y = "estimate", colour = "Group")) +
  geom_boxplot(outlier.shape = NA, fill = NA) +
  geom_point(position = position_jitter(0.15), size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.1, alpha = 0.5) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "DivNet: bias-corrected Shannon entropy",
       subtitle = "Error bars = 95% CI",
       x = NULL, y = "Shannon H′") +
  theme_bw() +
  theme(legend.position = "none")

print(p_shannon)
```

```{r}
#| label: divnet-hypothesis-test
# Test group differences using breakaway's betta on DivNet estimates
shannon_betta <- breakaway::betta(
  chats = shannon_dv$estimate,
  ses   = shannon_dv$std_error,
  X     = model.matrix(as.formula(paste("~", group_var)),
                        data = as.data.frame(sample_data(ps)))
)

cat("\n── DivNet + betta: Shannon diversity ~ ", group_var, " ──\n")
print(shannon_betta$table)
```

---

# 5. Beta Diversity

## 5.1 Distance Matrices

```{r}
#| label: compute-distances
dist_list <- list()

# Bray-Curtis (on relative abundance)
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
dist_list$bray <- phyloseq::distance(ps_rel, method = "bray")

# Weighted UniFrac
if (!is.null(phy_tree(ps, errorIfNULL = FALSE))) {
  dist_list$wunifrac <- phyloseq::distance(ps, method = "wunifrac")
  dist_list$unifrac  <- phyloseq::distance(ps, method = "unifrac")
} else {
  message("No tree – skipping UniFrac distances.")
}

# Aitchison distance (CLR-transformed Euclidean)
pseudocount <- 0.5
otu_clr <- microbiome::transform(ps, transform = "clr")
dist_list$aitchison <- vegan::vegdist(t(otu_table(otu_clr)), method = "euclidean")

cat("Distances computed:", paste(names(dist_list), collapse = ", "), "\n")
```

## 5.2 Ordination: PCoA / NMDS

```{r}
#| label: pcoa-plots
#| fig-width: 12
#| fig-height: 9
ordination_plots <- lapply(names(dist_list), function(dist_name) {
  ord <- ordinate(ps, method = "PCoA", distance = dist_list[[dist_name]])
  plot_ordination(ps, ord, color = group_var) +
    geom_point(size = 3) +
    stat_ellipse(level = 0.95, linetype = "dashed") +
    scale_colour_brewer(palette = "Set2") +
    labs(title = paste("PCoA –", dist_name),
         colour = group_var) +
    theme_bw()
})

names(ordination_plots) <- names(dist_list)

lapply(ordination_plots, print)
```

```{r}
#| label: nmds-bray
#| fig-width: 7
#| fig-height: 5
nmds_ord <- ordinate(ps_rel, method = "NMDS", distance = "bray")
cat("NMDS stress:", nmds_ord$stress, "\n")

plot_ordination(ps_rel, nmds_ord, color = group_var) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  annotate("text", x = Inf, y = -Inf,
           label = sprintf("Stress = %.3f", nmds_ord$stress),
           hjust = 1.1, vjust = -0.5) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "NMDS – Bray-Curtis", colour = group_var) +
  theme_bw()
```

## 5.3 PERMANOVA

PERMANOVA (adonis2) tests whether groups differ in community composition.

```{r}
#| label: permanova
meta_df <- as.data.frame(sample_data(ps))

permanova_results <- lapply(names(dist_list), function(dist_name) {
  formula_str <- paste("dist_list[[dist_name]] ~", group_var)
  res <- vegan::adonis2(
    as.formula(formula_str),
    data         = meta_df,
    permutations = n_permutations,
    by           = "margin"
  )
  cat(sprintf("\n── PERMANOVA – %s ──\n", dist_name))
  print(res)
  res
})
names(permanova_results) <- names(dist_list)
```

## 5.4 PERMDISP (Homogeneity of Dispersion)

A significant PERMANOVA could reflect differences in **location** (composition)
or **spread** (variability). PERMDISP tests whether groups have equal dispersion.

```{r}
#| label: permdisp
for (dist_name in names(dist_list)) {
  bd <- betadisper(dist_list[[dist_name]],
                    group = meta_df[[group_var]])
  disp_test <- permutest(bd, permutations = n_permutations)
  cat(sprintf("\n── PERMDISP – %s ──\n", dist_name))
  print(disp_test)
  plot(bd, main = paste("PERMDISP –", dist_name))
}
```

## 5.5 Summary Table

```{r}
#| label: permanova-summary-table
perm_summary <- do.call(rbind, lapply(names(permanova_results), function(dist_name) {
  res <- permanova_results[[dist_name]]
  data.frame(
    Distance = dist_name,
    R2       = round(res$R2[1], 4),
    F_stat   = round(res$F[1], 4),
    p_value  = res$`Pr(>F)`[1],
    stringsAsFactors = FALSE
  )
}))

perm_summary$p_adj <- p.adjust(perm_summary$p_value, method = alpha_p_adjust)
knitr::kable(perm_summary, caption = "PERMANOVA summary across distance metrics")
```

---

# 6. Save Results

```{r}
#| label: save-results
# Alpha diversity tables
write.csv(alpha_std,   file.path(results_dir, "alpha_standard_rarefied.csv"),  row.names = FALSE)
write.csv(ba_summary,  file.path(results_dir, "alpha_breakaway_richness.csv"), row.names = FALSE)
write.csv(shannon_dv,  file.path(results_dir, "alpha_divnet_shannon.csv"),     row.names = FALSE)

# Beta diversity tables
write.csv(perm_summary, file.path(results_dir, "beta_permanova_summary.csv"), row.names = FALSE)

# Save distance matrices
for (dist_name in names(dist_list)) {
  write.csv(as.matrix(dist_list[[dist_name]]),
            file.path(results_dir, paste0("dist_", dist_name, ".csv")))
}

message("Diversity results saved to: ", results_dir)
```

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
