---
title: "07 – Differential Abundance Analysis"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 4
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: false
  fig-width: 9
  fig-height: 6
---

# Overview

This notebook identifies taxa that are differentially abundant between groups
using two complementary compositional approaches:

1. **ANCOM-BC2** (Lin & Peddada 2020; Cao *et al.* 2023) – a log-linear model
   with bias correction that accounts for the compositional nature of
   microbiome data and provides FDR control. This is the primary method.

2. **Dirichlet regression** (`DirichletReg`) – models count proportions
   directly under the Dirichlet distribution. Useful as a confirmatory
   approach for taxa significant in ANCOM-BC2.

::: callout-note
DESeq2 is intentionally excluded because its negative binomial model treats
library sizes as absolute scales. For compositional data, ANCOM-BC2 is more
appropriate and less prone to false positives.
:::

---

# 1. User Parameters

```{r}
#| label: params
# ── Paths ─────────────────────────────────────────────────────────────────────
phyloseq_dir   <- here::here("data", "phyloseq")
ps_path        <- file.path(phyloseq_dir, "ps.rds")
da_results_dir <- here::here("data", "differential_abundance")

# ── Study design ──────────────────────────────────────────────────────────────
group_var     <- "Group"     # Primary grouping variable
reference_lvl <- NULL        # Reference level for pairwise comparisons.
                              # NULL = first alphabetical level

# Additional fixed effects (e.g., batch, sex) – set to NULL if not needed
fixed_effects <- NULL        # e.g., c("Batch", "Sex")

# ── Taxonomic level for analysis ──────────────────────────────────────────────
# For ANCOM-BC2: run on ASV level; we also repeat on Genus & Family
tax_levels <- c("ASV", "Genus", "Family")

# ── Minimum prevalence filter (before DA testing) ─────────────────────────────
# Keep only taxa present in >= this fraction of samples
prev_threshold <- 0.10   # 10% of samples

# ── ANCOM-BC2 settings ────────────────────────────────────────────────────────
ancombc2_alpha  <- 0.05   # FDR threshold
ancombc2_n_boot <- 1000   # Bootstrap iterations for sensitivity analysis
                            # (set to 0 to skip sensitivity, much faster)

# ── Dirichlet regression settings ─────────────────────────────────────────────
# Collapse to genus level for Dirichlet (ASV level is too sparse)
dirichlet_tax_level <- "Genus"
dirichlet_top_n     <- 40    # Top N genera by mean abundance to include
dirichlet_alpha     <- 0.05
```

---

# 2. Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(phyloseq)
  library(ANCOMBC)
  library(DirichletReg)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(patchwork)
  library(here)
})

dir.create(da_results_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# 3. Load and Prepare Data

```{r}
#| label: load-ps
ps <- readRDS(ps_path)
ps
```

```{r}
#| label: set-reference
if (!is.null(reference_lvl)) {
  meta <- as.data.frame(sample_data(ps))
  meta[[group_var]] <- relevel(factor(meta[[group_var]]), ref = reference_lvl)
  sample_data(ps) <- sample_data(meta)
  cat("Reference level set to:", reference_lvl, "\n")
} else {
  cat("Reference level: first alphabetical (",
      sort(unique(as.data.frame(sample_data(ps))[[group_var]]))[1], ")\n")
}
```

---

# 4. ANCOM-BC2

ANCOM-BC2 (Analysis of Compositions of Microbiomes with Bias Correction 2) is
a log-linear model that:

- Estimates and corrects for **sampling fractions** (library-size bias)
- Controls FDR via multiple testing correction
- Provides **sensitivity analysis** (pattern-based) to flag robust findings
- Supports pairwise comparisons, global tests, and continuous outcomes

Reference: Cao Y *et al.* (2023) *F1000Research* 13:369.

## 4.1 ASV-level ANCOM-BC2

```{r}
#| label: ancombc2-asvs
# Build formula
formula_str <- group_var
if (!is.null(fixed_effects)) {
  formula_str <- paste(c(formula_str, fixed_effects), collapse = " + ")
}
cat("ANCOM-BC2 formula: ~", formula_str, "\n")

# Prevalence filter
ps_prev <- phyloseq::filter_taxa(
  ps,
  function(x) mean(x > 0) >= prev_threshold,
  prune = TRUE
)
cat(sprintf("ASVs after prevalence filter (>= %.0f%%): %d\n",
            100 * prev_threshold, ntaxa(ps_prev)))

message("Running ANCOM-BC2 at ASV level...")

ancom_asv <- ANCOMBC::ancombc2(
  data          = ps_prev,
  assay_name    = "counts",
  tax_level     = NULL,   # NULL = use taxa as-is (ASV level)
  fix_formula   = formula_str,
  rand_formula  = NULL,
  p_adj_method  = "BH",
  prv_cut       = 0,
  lib_cut       = 0,
  s0_perc       = 0.05,
  group         = group_var,
  struc_zero    = TRUE,
  neg_lb        = TRUE,
  alpha         = ancombc2_alpha,
  n_cl          = 1,
  verbose       = TRUE,
  global        = TRUE,
  pairwise      = TRUE,
  dunnet        = FALSE,
  trend         = FALSE,
  iter_control  = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control    = list(tol = 1e-5, max_iter = 100),
  lme_control   = NULL,
  mdfdr_control = list(fwer_ctrl_method = "BH", B = ancombc2_n_boot),
  trend_control = list(contrast = NULL, node = NULL, solver = "ECOS", B = 100)
)
```

```{r}
#| label: ancombc2-asv-results
res_asv <- ancom_asv$res
cat(sprintf("\nSignificant ASVs (q < %.2f):\n", ancombc2_alpha))

sig_cols <- grep("^diff_", colnames(res_asv), value = TRUE)
for (col in sig_cols) {
  n_sig <- sum(res_asv[[col]], na.rm = TRUE)
  cat(sprintf("  %s: %d significant\n", sub("diff_", "", col), n_sig))
}
```

## 4.2 Genus-level ANCOM-BC2

Running ANCOM-BC2 at genus level provides a higher-level summary and increases
power by aggregating sparse ASV counts.

```{r}
#| label: ancombc2-genus
message("Running ANCOM-BC2 at Genus level...")

ancom_genus <- ANCOMBC::ancombc2(
  data         = ps_prev,
  assay_name   = "counts",
  tax_level    = "Genus",
  fix_formula  = formula_str,
  p_adj_method = "BH",
  prv_cut      = 0,
  lib_cut      = 0,
  s0_perc      = 0.05,
  group        = group_var,
  struc_zero   = TRUE,
  neg_lb       = TRUE,
  alpha        = ancombc2_alpha,
  n_cl         = 1,
  verbose      = FALSE,
  global       = TRUE,
  pairwise     = TRUE,
  dunnet       = FALSE,
  trend        = FALSE,
  iter_control = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control   = list(tol = 1e-5, max_iter = 100)
)

res_genus <- ancom_genus$res
cat(sprintf("\nGenera tested: %d\n", nrow(res_genus)))

sig_genus_cols <- grep("^diff_", colnames(res_genus), value = TRUE)
for (col in sig_genus_cols) {
  n_sig <- sum(res_genus[[col]], na.rm = TRUE)
  cat(sprintf("  %s: %d significant\n", sub("diff_", "", col), n_sig))
}
```

## 4.3 Visualise ANCOM-BC2 Results

### Volcano Plot (Genus level)

```{r}
#| label: ancombc2-volcano
#| fig-width: 8
#| fig-height: 6
lfc_col  <- grep("^lfc_",  colnames(res_genus), value = TRUE)[1]
se_col   <- grep("^se_",   colnames(res_genus), value = TRUE)[1]
q_col    <- grep("^q_",    colnames(res_genus), value = TRUE)[1]
diff_col <- grep("^diff_", colnames(res_genus), value = TRUE)[1]

if (!is.na(lfc_col)) {
  volcano_df <- res_genus %>%
    select(taxon, all_of(c(lfc_col, se_col, q_col, diff_col))) %>%
    setNames(c("Taxon", "LFC", "SE", "q", "Significant")) %>%
    mutate(
      q         = replace_na(q, 1),
      Direction = case_when(
        Significant & LFC > 0 ~ "Enriched",
        Significant & LFC < 0 ~ "Depleted",
        TRUE                  ~ "NS"
      )
    )

  top_labels <- volcano_df %>%
    filter(Significant) %>%
    arrange(q) %>%
    head(15)

  ggplot(volcano_df, aes(x = LFC, y = -log10(q), colour = Direction)) +
    geom_point(alpha = 0.6, size = 2.5) +
    geom_hline(yintercept = -log10(ancombc2_alpha),
               linetype = "dashed", colour = "grey40") +
    geom_vline(xintercept = 0, linetype = "solid", colour = "grey40") +
    ggrepel::geom_text_repel(data = top_labels,
                              aes(label = Taxon), size = 3, max.overlaps = 20) +
    scale_colour_manual(values = c(Enriched = "#E64B35", Depleted = "#4DBBD5",
                                    NS = "grey70")) +
    labs(
      title    = "ANCOM-BC2 – Volcano plot (Genus level)",
      subtitle = sub("lfc_", "", lfc_col),
      x        = "Log2 fold change",
      y        = expression(-log[10](q)),
      colour   = NULL
    ) +
    theme_bw()
}
```

### Heatmap of Top Significant Genera

```{r}
#| label: ancombc2-heatmap
#| fig-width: 9
#| fig-height: 7
if (!is.na(diff_col)) {
  sig_genera <- res_genus$taxon[res_genus[[diff_col]] == TRUE]

  if (length(sig_genera) > 0) {
    lfc_plot_df <- res_genus %>%
      filter(taxon %in% sig_genera) %>%
      select(taxon, all_of(grep("^lfc_", colnames(res_genus), value = TRUE)))

    lfc_long <- pivot_longer(lfc_plot_df,
                              cols      = -taxon,
                              names_to  = "Contrast",
                              values_to = "LFC") %>%
      mutate(Contrast = sub("lfc_", "", Contrast))

    ggplot(lfc_long, aes(x = Contrast, y = taxon, fill = LFC)) +
      geom_tile(colour = "white") +
      scale_fill_gradient2(low = "#4DBBD5", mid = "white", high = "#E64B35",
                            midpoint = 0) +
      labs(title = "ANCOM-BC2 – Log fold change heatmap (significant genera)",
           x = "Contrast", y = "Genus", fill = "LFC") +
      theme_bw() +
      theme(axis.text.y = element_text(size = 8),
            axis.text.x = element_text(angle = 30, hjust = 1))
  } else {
    message("No significant genera to display in heatmap.")
  }
}
```

### Bar Chart: Top Differential Genera

```{r}
#| label: ancombc2-barplot
#| fig-width: 8
#| fig-height: 6
if (!is.na(lfc_col)) {
  top_da_genera <- res_genus %>%
    filter(!is.na(.data[[q_col]])) %>%
    arrange(.data[[q_col]]) %>%
    head(20) %>%
    mutate(
      Direction = ifelse(.data[[lfc_col]] > 0, "Enriched", "Depleted"),
      Taxon     = reorder(taxon, .data[[lfc_col]])
    )

  ggplot(top_da_genera, aes(x = Taxon, y = .data[[lfc_col]],
                              fill = Direction)) +
    geom_col() +
    geom_errorbar(aes(ymin = .data[[lfc_col]] - 1.96 * .data[[se_col]],
                      ymax = .data[[lfc_col]] + 1.96 * .data[[se_col]]),
                  width = 0.3) +
    coord_flip() +
    scale_fill_manual(values = c(Enriched = "#E64B35", Depleted = "#4DBBD5")) +
    labs(title = "ANCOM-BC2 – Top 20 differential genera",
         subtitle = paste("Contrast:", sub("lfc_", "", lfc_col)),
         x = NULL, y = "Log2 fold change (± 95% CI)") +
    theme_bw() +
    theme(legend.position = "none")
}
```

## 4.4 ANCOM-BC2 Sensitivity Analysis

```{r}
#| label: ancombc2-sensitivity
if (!is.null(ancom_genus$res_global) && ancombc2_n_boot > 0) {
  cat("\nGlobal test results:\n")
  global_res <- ancom_genus$res_global
  cat(sprintf("Genera with significant global test: %d\n",
              sum(global_res$diff_abn, na.rm = TRUE)))
}

if (!is.null(ancom_genus$res_pair)) {
  cat("\nPairwise test summary:\n")
  pair_res <- ancom_genus$res_pair
  sig_pair_cols <- grep("^diff_", colnames(pair_res), value = TRUE)
  for (col in sig_pair_cols) {
    cat(sprintf("  %s: %d\n", sub("diff_", "", col),
                sum(pair_res[[col]], na.rm = TRUE)))
  }
}
```

---

# 5. Dirichlet Regression

Dirichlet regression models the proportional composition of the microbiome
under the Dirichlet distribution, naturally handling the compositional
constraint without log-ratio transformation.

We apply it at the genus level, using only the top genera to keep the model
tractable. It serves as a **confirmatory** method for ANCOM-BC2 findings.

## 5.1 Prepare Genus-level Data

```{r}
#| label: dirichlet-prep
# Aggregate to genus level
ps_genus_dr <- tax_glom(ps, "Genus")

# Select top N genera by mean abundance
top_genera  <- names(sort(taxa_sums(ps_genus_dr), decreasing = TRUE))[1:dirichlet_top_n]
ps_genus_dr <- prune_taxa(top_genera, ps_genus_dr)

# Convert to proportions (Dirichlet requires compositional data)
ps_prop <- transform_sample_counts(ps_genus_dr, function(x) x / sum(x))

otu_prop     <- as.data.frame(t(otu_table(ps_prop)))
colnames(otu_prop) <- paste0("g", seq_len(ncol(otu_prop)))
genus_labels <- as.vector(tax_table(ps_genus_dr)[, "Genus"])

meta_dr <- as.data.frame(sample_data(ps_genus_dr))
meta_dr[[group_var]] <- factor(meta_dr[[group_var]])

# DirichletReg needs a DirichletRegData object
dr_data <- cbind(meta_dr, otu_prop)
Y <- DirichletReg::DR_data(dr_data[, colnames(otu_prop)])

cat("Dirichlet input: ", nrow(dr_data), "samples ×",
    ncol(otu_prop), "genera\n")
```

## 5.2 Fit Dirichlet Regression Model

```{r}
#| label: dirichlet-fit
formula_dr <- as.formula(paste("Y ~", group_var))
if (!is.null(fixed_effects)) {
  formula_dr <- as.formula(paste("Y ~", paste(c(group_var, fixed_effects), collapse = " + ")))
}

message("Fitting Dirichlet regression (this may take a few minutes)...")

dr_model <- DirichletReg::DirichReg(
  formula_dr,
  data  = dr_data,
  model = "alternative"  # "alternative" parameterisation for covariate effects
)

summary_dr <- summary(dr_model)
```

## 5.3 Extract Significant Associations

```{r}
#| label: dirichlet-results
coef_table    <- summary_dr$coef.mat
group_levels  <- levels(meta_dr[[group_var]])[-1]
group_pattern <- paste(group_var, group_levels, sep = "", collapse = "|")

dr_results <- do.call(rbind, lapply(seq_along(genus_labels), function(i) {
  genus     <- genus_labels[i]
  all_rows  <- rownames(coef_table)
  relevant_rows <- grep(group_pattern, all_rows)
  if (length(relevant_rows) == 0) return(NULL)

  row_data <- coef_table[relevant_rows, , drop = FALSE]

  do.call(rbind, lapply(seq_len(nrow(row_data)), function(j) {
    data.frame(
      Genus    = genus,
      Contrast = rownames(row_data)[j],
      Estimate = row_data[j, "Estimate"],
      SE       = row_data[j, "Std. Error"],
      z        = row_data[j, "z value"],
      p_value  = row_data[j, "Pr(>|z|)"],
      stringsAsFactors = FALSE
    )
  }))
}))

dr_results <- dr_results %>%
  mutate(q_value = p.adjust(p_value, method = "BH")) %>%
  arrange(q_value)

cat(sprintf("\nSignificant genera (Dirichlet, q < %.2f): %d\n",
            dirichlet_alpha,
            sum(dr_results$q_value < dirichlet_alpha, na.rm = TRUE)))

knitr::kable(
  filter(dr_results, q_value < dirichlet_alpha),
  digits  = 4,
  caption = paste("Dirichlet regression – significant genera (q <",
                  dirichlet_alpha, ")")
)
```

```{r}
#| label: dirichlet-volcano
#| fig-width: 7
#| fig-height: 5
ggplot(dr_results, aes(x = Estimate, y = -log10(q_value),
                        colour = q_value < dirichlet_alpha)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = -log10(dirichlet_alpha),
             linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#E64B35"),
                      labels = c("NS", "Significant"), name = NULL) +
  ggrepel::geom_text_repel(
    data = filter(dr_results, q_value < dirichlet_alpha),
    aes(label = Genus), size = 3, max.overlaps = 15
  ) +
  labs(title = "Dirichlet regression – Volcano plot (Genus level)",
       x = "Coefficient estimate", y = expression(-log[10](q))) +
  theme_bw()
```

---

# 6. Comparison: ANCOM-BC2 vs Dirichlet Regression

```{r}
#| label: compare-methods
sig_ancombc2 <- if (!is.na(diff_col)) {
  res_genus$taxon[res_genus[[diff_col]] == TRUE & !is.na(res_genus[[diff_col]])]
} else {
  character(0)
}

sig_dirichlet <- dr_results$Genus[dr_results$q_value < dirichlet_alpha &
                                     !is.na(dr_results$q_value)]

both   <- intersect(sig_ancombc2, sig_dirichlet)
only_a <- setdiff(sig_ancombc2, sig_dirichlet)
only_d <- setdiff(sig_dirichlet, sig_ancombc2)

cat("── Method concordance ──────────────────────────────\n")
cat(sprintf("  Significant in ANCOM-BC2 only:      %d genera\n", length(only_a)))
cat(sprintf("  Significant in Dirichlet only:      %d genera\n", length(only_d)))
cat(sprintf("  Significant in BOTH methods:        %d genera\n", length(both)))
cat("────────────────────────────────────────────────────\n")

if (length(both) > 0) {
  cat("Genera significant in both methods:\n  ",
      paste(both, collapse = ", "), "\n")
}
```

---

# 7. Save Results

```{r}
#| label: save-results
# ANCOM-BC2 results
write.csv(res_asv,   file.path(da_results_dir, "ancombc2_asv.csv"),   row.names = FALSE)
write.csv(res_genus, file.path(da_results_dir, "ancombc2_genus.csv"), row.names = FALSE)

# Dirichlet results
write.csv(dr_results, file.path(da_results_dir, "dirichlet_genus.csv"), row.names = FALSE)

# Concordance summary
concordance <- data.frame(
  Genus     = unique(c(sig_ancombc2, sig_dirichlet)),
  ANCOMBC2  = unique(c(sig_ancombc2, sig_dirichlet)) %in% sig_ancombc2,
  Dirichlet = unique(c(sig_ancombc2, sig_dirichlet)) %in% sig_dirichlet
)
concordance$Both <- concordance$ANCOMBC2 & concordance$Dirichlet
write.csv(concordance, file.path(da_results_dir, "concordance_summary.csv"), row.names = FALSE)

# Save ANCOM-BC2 model objects
saveRDS(ancom_asv,   file.path(da_results_dir, "ancombc2_asv_model.rds"))
saveRDS(ancom_genus, file.path(da_results_dir, "ancombc2_genus_model.rds"))

message("Differential abundance results saved to: ", da_results_dir)
```

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
