---
title: "00 – Environment Setup & Package Installation"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  eval: false   # Set to true only when actively installing
  message: true
  warning: true
---

# Overview

This notebook installs all R packages required by the 16S pipeline and records
the session environment for reproducibility. Run this **once** before executing
the analysis notebooks.

::: callout-note
Set `eval: true` in the `execute` block above (or run each chunk manually)
when you actually want to install packages. Leaving `eval: false` prevents
accidental re-installation on every render.
:::

---

# 1. CRAN Packages

```{r}
#| label: install-cran
cran_pkgs <- c(
  # Core workflow
  "BiocManager",    # Bioconductor installer
  "remotes",        # GitHub installs
  "rmarkdown",      # Report rendering
  "knitr",          # Knitting

  # Data wrangling & plotting
  "tidyverse",      # dplyr, tidyr, ggplot2, readr, purrr, forcats, stringr
  "patchwork",      # Compose ggplot panels
  "scales",         # Axis formatting
  "ggrepel",        # Non-overlapping labels
  "pheatmap",       # Heatmaps
  "RColorBrewer",   # Colour palettes
  "viridis",        # Perceptually uniform palettes

  # Compositional statistics
  "DirichletReg",   # Dirichlet regression for differential abundance
  "compositions",   # CoDa tools (CLR, ILR)

  # Phylogenetics
  "ape",            # Phylogenetic tree handling
  "phangorn",       # Tree construction (NJ + GTR)

  # Diversity
  "vegan",          # Community ecology (PERMANOVA, NMDS, etc.)

  # Utilities
  "jsonlite",       # JSON I/O (rFigaro dependency)
  "parallel",       # Parallel processing
  "here"            # Relative paths
)

installed <- rownames(installed.packages())
to_install <- setdiff(cran_pkgs, installed)

if (length(to_install) > 0) {
  message("Installing from CRAN: ", paste(to_install, collapse = ", "))
  install.packages(to_install, repos = "https://cloud.r-project.org")
} else {
  message("All CRAN packages already installed.")
}
```

---

# 2. Bioconductor Packages

```{r}
#| label: install-bioc
bioc_pkgs <- c(
  # 16S / amplicon
  "dada2",          # ASV denoising (core pipeline step)
  "DECIPHER",       # Taxonomic classification (IdTaxa)
  "decontam",       # Contaminant identification
  "phyloseq",       # Microbiome data container & ecology
  "microbiome",     # phyloseq extensions and utilities

  # Differential abundance
  "ANCOMBC",        # ANCOM-BC2 – compositionally aware DA testing

  # Diversity with bias correction
  "breakaway",      # Bias-corrected richness & betta() regression
  "DivNet",         # Network-aware diversity estimation (Shannon, Simpson)

  # Quality & short reads
  "ShortRead",      # FASTQ I/O (also used by rFigaro optionally)
  "Biostrings"      # DNA string manipulation
)

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

installed_bioc <- rownames(installed.packages())
to_install_bioc <- setdiff(bioc_pkgs, installed_bioc)

if (length(to_install_bioc) > 0) {
  message("Installing from Bioconductor: ", paste(to_install_bioc, collapse = ", "))
  BiocManager::install(to_install_bioc, ask = FALSE)
} else {
  message("All Bioconductor packages already installed.")
}
```

---

# 3. rFigaro (Local)

The `rFigaro` package is included as a local source in `rFigaro/`. Install it
once from source:

```{r}
#| label: install-rfigaro
rfigaro_path <- here::here("rFigaro")

if (!requireNamespace("rFigaro", quietly = TRUE)) {
  message("Installing rFigaro from local source: ", rfigaro_path)
  remotes::install_local(rfigaro_path, upgrade = "never")
} else {
  message("rFigaro already installed. Version: ",
          packageVersion("rFigaro"))
}
```

To force reinstall (e.g., after updating the package source):

```{r}
#| label: reinstall-rfigaro
remotes::install_local(here::here("rFigaro"), force = TRUE, upgrade = "never")
```

---

# 4. External Tool: cutadapt

Primer trimming uses **cutadapt** (Python), called from R via `system2()`.
Install it outside R:

```{bash}
#| label: install-cutadapt
# Option A – conda/mamba (recommended)
mamba install -c bioconda cutadapt

# Option B – pip
pip install cutadapt

# Verify
cutadapt --version
```

Verify that R can find it:

```{r}
#| label: check-cutadapt
#| eval: true
cutadapt_check <- system2("cutadapt", "--version", stdout = TRUE, stderr = TRUE)
if (length(cutadapt_check) > 0) {
  cat("cutadapt found:", cutadapt_check[1], "\n")
} else {
  warning("cutadapt not found in PATH. Set `cutadapt_path` manually in notebook 01.")
}
```

---

# 5. Verify Installation

```{r}
#| label: verify
#| eval: true
pkgs_to_check <- c(
  "dada2", "phyloseq", "DECIPHER", "decontam",
  "ANCOMBC", "breakaway", "DivNet",
  "vegan", "DirichletReg", "microbiome",
  "ape", "phangorn", "ShortRead",
  "ggplot2", "dplyr", "tidyr", "patchwork"
)

status <- sapply(pkgs_to_check, function(p) {
  tryCatch({
    requireNamespace(p, quietly = TRUE)
    as.character(packageVersion(p))
  }, error = function(e) "NOT INSTALLED")
})

df <- data.frame(
  Package = names(status),
  Version = unname(status),
  Status  = ifelse(status == "NOT INSTALLED", "MISSING", "OK"),
  stringsAsFactors = FALSE
)

knitr::kable(df, caption = "Package installation status")
```

```{r}
#| label: verify-rfigaro
#| eval: true
rfig_ok <- requireNamespace("rFigaro", quietly = TRUE)
cat("rFigaro installed:", rfig_ok, "\n")
if (rfig_ok) cat("rFigaro version:", as.character(packageVersion("rFigaro")), "\n")
```

---

# 6. Session Information

```{r}
#| label: session-info
#| eval: true
sessionInfo()
```
