---
title: "01 – Primer Trimming"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: false
---

# Overview

This notebook removes primer sequences from raw paired-end FASTQ files using
**cutadapt**. Only reads in which both primers are found (and removed) are
retained; spurious primer-less reads are discarded.

The workflow:

1. Locate raw FASTQ files
2. Inspect a quality profile of raw reads
3. Run cutadapt on each sample (parallelised)
4. Verify trimming success

---

# 1. User Parameters

::: callout-important
Edit this section before running the notebook.
:::

```{r}
#| label: params
# ── Paths ─────────────────────────────────────────────────────────────────────
raw_dir     <- here::here("data", "raw")         # Directory with raw FASTQs
trimmed_dir <- here::here("data", "trimmed")     # Output for trimmed FASTQs
cutadapt_path <- "cutadapt"  # Full path if not in PATH, e.g. "/usr/bin/cutadapt"

# ── Primers (5' → 3', without ambiguity codes unless intentional) ─────────────
# 515F / 806R (Parada et al. 2016 / Apprill et al. 2015) – V4 region example
primer_fwd <- "GTGYCAGCMGCCGCGGTAA"   # Forward primer
primer_rev <- "GGACTACNVGGGTWTCTAAT" # Reverse primer

# ── Cutadapt settings ─────────────────────────────────────────────────────────
min_length        <- 50     # Discard reads shorter than this after trimming (bp)
error_rate        <- 0.1    # Maximum error rate for primer matching (0–1)
discard_untrimmed <- TRUE   # Discard reads where the primer was not found
n_cores           <- 4      # Parallel jobs (set to 1 if issues arise)

# ── FASTQ naming convention ────────────────────────────────────────────────────
# Pattern to identify forward and reverse reads in filename
fwd_pattern <- "_R1"   # e.g. sample1_R1.fastq.gz
rev_pattern <- "_R2"
```

---

# 2. Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(dada2)
  library(ShortRead)
  library(ggplot2)
  library(dplyr)
  library(here)
})
```

```{r}
#| label: create-dirs
dir.create(trimmed_dir, recursive = TRUE, showWarnings = FALSE)
message("Trimmed FASTQ output directory: ", trimmed_dir)
```

```{r}
#| label: check-cutadapt
ca_version <- tryCatch(
  system2(cutadapt_path, "--version", stdout = TRUE, stderr = TRUE),
  error = function(e) NULL
)

if (is.null(ca_version) || length(ca_version) == 0) {
  stop(
    "cutadapt not found at: ", cutadapt_path,
    "\nInstall cutadapt (see docs/00_setup.qmd) or update `cutadapt_path`."
  )
}
cat("cutadapt version:", ca_version[1], "\n")
```

---

# 3. Locate Raw FASTQ Files

```{r}
#| label: list-fastqs
fwd_files <- sort(list.files(raw_dir, pattern = paste0(fwd_pattern, ".*\\.fastq(\\.gz)?$"),
                              full.names = TRUE))
rev_files <- sort(list.files(raw_dir, pattern = paste0(rev_pattern, ".*\\.fastq(\\.gz)?$"),
                              full.names = TRUE))

if (length(fwd_files) == 0) stop("No forward FASTQ files found in: ", raw_dir)
if (length(fwd_files) != length(rev_files)) {
  stop("Unequal number of forward (", length(fwd_files),
       ") and reverse (", length(rev_files), ") files.")
}

# Extract sample names (everything before the R1/R2 pattern)
sample_names <- sub(paste0(fwd_pattern, ".*"), "", basename(fwd_files))
# Remove trailing separator characters
sample_names <- gsub("[._-]$", "", sample_names)

cat("Samples found:", length(sample_names), "\n")
head(data.frame(Sample = sample_names,
                Forward = basename(fwd_files),
                Reverse = basename(rev_files)), 10)
```

---

# 4. Raw Quality Profiles

Visualise quality profiles for a random subset of samples before trimming.
This helps confirm primer presence and assess overall read quality.

```{r}
#| label: raw-quality
#| fig-width: 10
#| fig-height: 5
n_plot <- min(4, length(fwd_files))
idx    <- sort(sample(length(fwd_files), n_plot))

p_fwd <- plotQualityProfile(fwd_files[idx]) +
  ggtitle("Raw forward reads – quality profiles") +
  theme_bw()

p_rev <- plotQualityProfile(rev_files[idx]) +
  ggtitle("Raw reverse reads – quality profiles") +
  theme_bw()

print(p_fwd)
print(p_rev)
```

---

# 5. Check Primer Orientation

Primers can appear in multiple orientations. We check all four orientations
(forward, complement, reverse complement, reverse) to ensure we capture them all.

```{r}
#| label: primer-utils
# Orientation helper functions
rc <- function(seq) as.character(Biostrings::reverseComplement(Biostrings::DNAString(seq)))

primer_orientations <- function(primer) {
  c(
    Forward    = primer,
    Complement = as.character(Biostrings::complement(Biostrings::DNAString(primer))),
    RevComp    = rc(primer),
    Reverse    = paste(rev(strsplit(primer, "")[[1]]), collapse = "")
  )
}

fwd_orients <- primer_orientations(primer_fwd)
rev_orients <- primer_orientations(primer_rev)

cat("Forward primer orientations:\n"); print(fwd_orients)
cat("\nReverse primer orientations:\n"); print(rev_orients)
```

```{r}
#| label: count-primers
count_primers <- function(primer, file_path) {
  hits <- vcountPattern(primer,
                        sread(readFastq(file_path)),
                        fixed = FALSE)
  sum(hits > 0)
}

# Count occurrences in first sample only (as a check)
cat("\nPrimer counts in first sample forward read:\n")
sapply(fwd_orients, count_primers, file_path = fwd_files[1])

cat("\nPrimer counts in first sample reverse read:\n")
sapply(rev_orients, count_primers, file_path = rev_files[1])
```

---

# 6. Run Cutadapt

Cutadapt is run for each sample. We trim:

- The forward primer from the 5' end of R1 reads
- The reverse complement of the reverse primer from the 3' end of R1 reads
- The reverse primer from the 5' end of R2 reads
- The reverse complement of the forward primer from the 3' end of R2 reads

```{r}
#| label: define-outputs
fwd_trimmed <- file.path(trimmed_dir, basename(fwd_files))
rev_trimmed <- file.path(trimmed_dir, basename(rev_files))
```

```{r}
#| label: run-cutadapt
# Build cutadapt flags
flags_base <- c(
  "-g", primer_fwd,         # 5' adapter on R1
  "-a", rc(primer_rev),     # 3' adapter on R1 (RC of rev primer)
  "-G", primer_rev,         # 5' adapter on R2
  "-A", rc(primer_fwd),     # 3' adapter on R2 (RC of fwd primer)
  "-m", min_length,
  "--error-rate", error_rate,
  if (discard_untrimmed) "--discard-untrimmed",
  "-j", n_cores             # parallel
)

cutadapt_log <- vector("list", length(sample_names))
names(cutadapt_log) <- sample_names

for (i in seq_along(sample_names)) {
  flags <- c(
    flags_base,
    "-o", fwd_trimmed[i],
    "-p", rev_trimmed[i],
    fwd_files[i],
    rev_files[i]
  )

  log_out <- system2(cutadapt_path, args = flags,
                     stdout = TRUE, stderr = TRUE)
  cutadapt_log[[i]] <- log_out

  # Parse read retention from cutadapt output
  pairs_in  <- grep("Total read pairs processed", log_out, value = TRUE)
  pairs_out <- grep("Pairs written", log_out, value = TRUE)
  cat(sprintf("[%d/%d] %s | %s | %s\n",
              i, length(sample_names), sample_names[i],
              ifelse(length(pairs_in)  > 0, trimws(pairs_in[1]),  "?"),
              ifelse(length(pairs_out) > 0, trimws(pairs_out[1]), "?")))
}

message("Primer trimming complete.")
```

---

# 7. Trimming Summary

```{r}
#| label: trimming-summary
parse_cutadapt_log <- function(log_lines, sample) {
  extract_n <- function(pattern) {
    line <- grep(pattern, log_lines, value = TRUE)[1]
    if (is.na(line)) return(NA_integer_)
    as.integer(gsub("[^0-9]", "", regmatches(line, regexpr("[0-9,]+", line))))
  }
  data.frame(
    Sample      = sample,
    Pairs_in    = extract_n("Total read pairs processed"),
    Pairs_out   = extract_n("Pairs written"),
    stringsAsFactors = FALSE
  )
}

summary_df <- do.call(rbind, Map(parse_cutadapt_log,
                                  cutadapt_log, sample_names))
summary_df <- summary_df %>%
  dplyr::mutate(
    Retention_pct = round(Pairs_out / Pairs_in * 100, 1)
  )

knitr::kable(summary_df, caption = "Cutadapt read retention per sample")
```

```{r}
#| label: retention-plot
#| fig-width: 8
#| fig-height: 4
ggplot(summary_df, aes(x = reorder(Sample, Retention_pct),
                        y = Retention_pct,
                        fill = Retention_pct < 70)) +
  geom_col() +
  geom_hline(yintercept = 70, linetype = "dashed", colour = "red") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
                    guide  = "none") +
  coord_flip() +
  labs(title = "Read retention after primer trimming",
       x = NULL, y = "Pairs retained (%)") +
  theme_bw()
```

::: callout-warning
Samples with < 70% retention (red) may have sequencing issues or an incorrect
primer specification. Check the cutadapt logs.
:::

---

# 8. Post-trimming Quality Profiles

```{r}
#| label: trimmed-quality
#| fig-width: 10
#| fig-height: 5
p_fwd_t <- plotQualityProfile(fwd_trimmed[idx]) +
  ggtitle("Trimmed forward reads – quality profiles") +
  theme_bw()

p_rev_t <- plotQualityProfile(rev_trimmed[idx]) +
  ggtitle("Trimmed reverse reads – quality profiles") +
  theme_bw()

print(p_fwd_t)
print(p_rev_t)
```

---

# 9. Save Sample Manifest

```{r}
#| label: save-manifest
manifest <- data.frame(
  Sample   = sample_names,
  Fwd_raw  = fwd_files,
  Rev_raw  = rev_files,
  Fwd_trim = fwd_trimmed,
  Rev_trim = rev_trimmed,
  stringsAsFactors = FALSE
)

manifest_path <- here::here("data", "trimmed", "sample_manifest.csv")
write.csv(manifest, manifest_path, row.names = FALSE)
message("Sample manifest saved to: ", manifest_path)
```

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
