---
title: "02 – Trim Parameter Optimisation with rFigaro"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: false
---

# Overview

This notebook uses **rFigaro** – an R implementation of the FIGARO algorithm
(Weinstein *et al.* 2019) – to determine optimal **truncation lengths**
(`truncLen`) and **maximum expected errors** (`maxEE`) for DADA2.

FIGARO scores every possible trim position combination by modelling the
cumulative expected error across reads. The top candidates balance:

1. **Read retention** (% of read pairs that survive the `maxEE` filter)
2. **Overlap** (trimmed reads must still overlap by ≥ `min_overlap` bp)

::: callout-note
**Input:** Primer-trimmed FASTQs from notebook `01`.
**Output:** Ranked `data.frame` of trim parameter candidates + JSON file.
:::

---

# 1. User Parameters

::: callout-important
Edit this section before running.
:::

```{r}
#| label: params
# ── Paths ─────────────────────────────────────────────────────────────────────
trimmed_dir  <- here::here("data", "trimmed")
figaro_dir   <- here::here("data", "figaro")   # rFigaro output directory

# ── Amplicon biology ──────────────────────────────────────────────────────────
# Length of the amplicon WITHOUT primers (bp).
# V4 region (515F/806R): ~253 bp; V3-V4 (~460 bp); V1-V3 (~490 bp)
amplicon_length <- 253

# Primer lengths (already trimmed, so set to 0 unless primers are still present)
# If running rFigaro on TRIMMED reads (recommended), set both to 0.
# If running on RAW reads, specify actual primer lengths.
forward_primer_length <- 0L
reverse_primer_length <- 0L

# ── FIGARO settings ───────────────────────────────────────────────────────────
min_overlap     <- 20L    # Minimum paired-read overlap for merging
percentile      <- 83L    # EE percentile for curve fitting (default 83)
n_cores         <- 4L     # Parallel cores
top_n           <- 20L    # Number of top candidates to report

# ── File naming ────────────────────────────────────────────────────────────────
naming_standard <- "nononsense"  # See rFigaro::figaro() docs for options
# "nononsense" = sample_R1.fastq.gz / sample_R2.fastq.gz
# "illumina"   = Illumina standard (sampleID_S*_L*_R{1,2}_001.fastq.gz)
```

---

# 2. Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(rFigaro)
  library(dada2)
  library(ggplot2)
  library(patchwork)
  library(here)
})

dir.create(figaro_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# 3. Quality Profiles (Reminder)

Visualise quality profiles of a few trimmed samples to guide your expectations.
The rFigaro output should suggest truncation points consistent with what you
see here.

```{r}
#| label: quality-profiles
#| fig-width: 10
#| fig-height: 5
fwd_files <- sort(list.files(trimmed_dir,
                              pattern = "_R1.*\\.fastq(\\.gz)?$",
                              full.names = TRUE))
rev_files <- sort(list.files(trimmed_dir,
                              pattern = "_R2.*\\.fastq(\\.gz)?$",
                              full.names = TRUE))

if (length(fwd_files) == 0) stop("No trimmed FASTQs found in: ", trimmed_dir)

n_plot <- min(4, length(fwd_files))
idx    <- sort(sample(length(fwd_files), n_plot))

p_fwd <- plotQualityProfile(fwd_files[idx]) +
  ggtitle("Trimmed forward reads") +
  theme_bw()
p_rev <- plotQualityProfile(rev_files[idx]) +
  ggtitle("Trimmed reverse reads") +
  theme_bw()

p_fwd / p_rev
```

---

# 4. Run rFigaro

```{r}
#| label: run-figaro
message("Running rFigaro on: ", trimmed_dir)
message("  Amplicon length:  ", amplicon_length, " bp")
message("  Minimum overlap:  ", min_overlap, " bp")
message("  EE percentile:    ", percentile)

figaro_result <- rFigaro::figaro(
  input_directory       = trimmed_dir,
  amplicon_length       = amplicon_length,
  forward_primer_length = forward_primer_length,
  reverse_primer_length = reverse_primer_length,
  min_overlap           = min_overlap,
  naming_standard       = naming_standard,
  percentile            = percentile,
  n_cores               = n_cores,
  generate_plots        = TRUE,
  output_directory      = figaro_dir,
  output_filename       = "trimParameters.json",
  top_n                 = top_n
)
```

---

# 5. Interpret Results

```{r}
#| label: print-results
print(figaro_result)
```

```{r}
#| label: results-table
params_df <- figaro_result$parameters

knitr::kable(
  params_df,
  digits  = 2,
  caption = "rFigaro trim parameter candidates (ranked by score)",
  col.names = c("Fwd truncLen", "Rev truncLen",
                "Fwd maxEE", "Rev maxEE",
                "Retention (%)", "Score")
)
```

---

# 6. Expected Error Curves

These curves show the cumulative expected error as a function of read position.
Truncation at a position just before quality deteriorates sharply maximises
both error filtering and read retention.

```{r}
#| label: ee-plots
#| fig-width: 10
#| fig-height: 5
if (!is.null(figaro_result$forward_curve$plot) &&
    !is.null(figaro_result$reverse_curve$plot)) {
  figaro_result$forward_curve$plot / figaro_result$reverse_curve$plot
} else {
  message("EE plots not available (ggplot2 may not be installed).")
}
```

---

# 7. Score vs Retention Trade-off

```{r}
#| label: score-plot
#| fig-width: 7
#| fig-height: 5
ggplot(params_df,
       aes(x = read_retention_percent,
           y = score,
           colour = forward_trim,
           shape  = factor(reverse_trim))) +
  geom_point(size = 3) +
  scale_colour_viridis_c(name = "Fwd truncLen") +
  labs(
    title   = "rFigaro: score vs read retention",
    x       = "Estimated read retention (%)",
    y       = "FIGARO score",
    shape   = "Rev truncLen",
    caption = "Higher score = better; top-left is ideal"
  ) +
  theme_bw()
```

---

# 8. Select & Record Parameters for DADA2

```{r}
#| label: select-params
# Default: use top-ranked candidate
best <- params_df[1, ]

# ── Override here if needed (e.g. prefer higher retention) ────────────────────
# best <- params_df[params_df$read_retention_percent >= 75, ][1, ]

cat("══════════════════════════════════════════\n")
cat("  Selected DADA2 parameters\n")
cat("══════════════════════════════════════════\n")
cat(sprintf("  truncLen  = c(%d, %d)\n", best$forward_trim, best$reverse_trim))
cat(sprintf("  maxEE     = c(%d, %d)\n", best$forward_max_ee, best$reverse_max_ee))
cat(sprintf("  Retention = %.1f%%\n",     best$read_retention_percent))
cat(sprintf("  Score     = %.3f\n",       best$score))
cat("══════════════════════════════════════════\n")
```

```{r}
#| label: save-params
# Save selected parameters for downstream use
dada2_params <- list(
  truncLen        = c(best$forward_trim, best$reverse_trim),
  maxEE           = c(best$forward_max_ee, best$reverse_max_ee),
  retention_pct   = best$read_retention_percent,
  figaro_score    = best$score,
  amplicon_length = amplicon_length,
  min_overlap     = min_overlap,
  timestamp       = Sys.time()
)

params_path <- here::here("data", "figaro", "dada2_params.rds")
saveRDS(dada2_params, params_path)
message("DADA2 parameters saved to: ", params_path)
```

---

# 9. Notes on Quality Score Schemes

::: callout-note
**This section is informational. Action required in notebook 03.**
:::

Modern Illumina platforms differ in how they encode quality scores:

| Platform | Q-score type | DADA2 implication |
|----------|-------------|-------------------|
| MiSeq, HiSeq | **Continuous** | Standard error model learning works well |
| NovaSeq 6000, NextSeq | **Binned** (2-channel) | Binned scores compress quality variation; use `LOESS` smoothing or `selfConsist = TRUE` |

rFigaro uses the raw Q-score distribution and is valid for **both** platforms.
For DADA2, if your data come from a **binned** platform, you will use
`loessErrfun` or enable `pool = TRUE` / `selfConsist = TRUE` as described in
notebook 03.

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
