---
title: "03 – DADA2 Processing"
subtitle: "16S Amplicon Sequencing Analysis Pipeline"
author: "Fundação GIMM – Gulbenkian Institute for Molecular Medicine"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 4
    theme: flatly
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: true
  warning: false
---

# Overview

This notebook runs the full **DADA2** pipeline (Callahan *et al.*, *Nature Methods*
2016) on primer-trimmed paired-end reads to generate an **ASV table** (Amplicon
Sequence Variants) and assigns taxonomy.

Steps:

1. Quality filtering & truncation
2. Error model learning *(with platform-aware handling of continuous vs binned Q-scores)*
3. Sample inference (denoising)
4. Paired-end read merging
5. Chimera removal
6. Read tracking
7. Taxonomy assignment (SILVA / GTDB)

---

# 1. User Parameters

::: callout-important
Edit this section before running.
:::

```{r}
#| label: params
# ── Paths ─────────────────────────────────────────────────────────────────────
trimmed_dir        <- here::here("data", "trimmed")
dada2_dir          <- here::here("data", "dada2")
figaro_params_path <- here::here("data", "figaro", "dada2_params.rds")

# ── Quality score platform ────────────────────────────────────────────────────
# Set to TRUE for NovaSeq 6000 / NextSeq (2-channel, binned Q-scores)
# Set to FALSE for MiSeq / HiSeq (continuous Q-scores)
binned_quality <- FALSE

# ── Trim parameters ────────────────────────────────────────────────────────────
# These are loaded from rFigaro output by default.
# Uncomment and edit to override manually:
# truncLen  <- c(240, 200)
# maxEE     <- c(2, 2)

# ── Additional filterAndTrim settings ─────────────────────────────────────────
truncQ        <- 2     # Truncate at first base with Q ≤ this value
rm_phiX       <- TRUE  # Remove PhiX spike-in reads
min_length    <- 50    # Minimum read length after truncation

# ── Merging settings ──────────────────────────────────────────────────────────
min_overlap   <- 20    # Minimum overlap for merging
max_mismatch  <- 0     # Maximum mismatches in overlap region (0 = exact)

# ── Chimera removal ────────────────────────────────────────────────────────────
chimera_method <- "consensus"  # "consensus" or "pooled"

# ── Taxonomy ──────────────────────────────────────────────────────────────────
# Path to SILVA reference database (download from https://zenodo.org/record/4587955)
# or GTDB (https://forum.qiime2.org/t/gtdb-taxonomy-for-dada2/23747)
taxonomy_db      <- "/path/to/silva_nr99_v138.1_train_set.fa.gz"
taxonomy_species <- "/path/to/silva_species_assignment_v138.1.fa.gz"

# ── Parallelism ────────────────────────────────────────────────────────────────
n_cores <- 4L  # Number of cores (use FALSE on Windows if multiprocessing fails)

# ── Pooling strategy ──────────────────────────────────────────────────────────
# "independent" (default, fastest), "pseudo" (better sensitivity), or TRUE (pooled)
pool_method <- "pseudo"
```

---

# 2. Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(dada2)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(patchwork)
  library(here)
})

dir.create(dada2_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(dada2_dir, "filtered"), recursive = TRUE, showWarnings = FALSE)

message("DADA2 version: ", packageVersion("dada2"))
```

---

# 3. Load rFigaro Parameters

```{r}
#| label: load-params
if (file.exists(figaro_params_path)) {
  figaro_params <- readRDS(figaro_params_path)
  message("Loaded rFigaro parameters from: ", figaro_params_path)

  # Use rFigaro values unless manually overridden above
  if (!exists("truncLen")) truncLen <- figaro_params$truncLen
  if (!exists("maxEE"))    maxEE    <- figaro_params$maxEE

  cat("  truncLen:", truncLen, "\n")
  cat("  maxEE:   ", maxEE,    "\n")
  cat("  Estimated read retention:", figaro_params$retention_pct, "%\n")
} else {
  warning("rFigaro parameter file not found. Using manually defined truncLen/maxEE.")
  if (!exists("truncLen") || !exists("maxEE")) {
    stop("Please define `truncLen` and `maxEE` in the User Parameters section.")
  }
}

cat("\n── Platform quality score scheme ──\n")
if (binned_quality) {
  cat("  Binned Q-scores (NovaSeq/NextSeq)\n")
  cat("  → Will use loessErrfun or selfConsist error modelling.\n")
} else {
  cat("  Continuous Q-scores (MiSeq/HiSeq)\n")
  cat("  → Standard DADA2 error modelling.\n")
}
```

---

# 4. Locate Trimmed FASTQs

```{r}
#| label: list-fastqs
fwd_trimmed <- sort(list.files(trimmed_dir,
                                pattern = "_R1.*\\.fastq(\\.gz)?$",
                                full.names = TRUE))
rev_trimmed <- sort(list.files(trimmed_dir,
                                pattern = "_R2.*\\.fastq(\\.gz)?$",
                                full.names = TRUE))

if (length(fwd_trimmed) == 0) stop("No trimmed FASTQs found in: ", trimmed_dir)
stopifnot(length(fwd_trimmed) == length(rev_trimmed))

sample_names <- sub("_R1.*", "", basename(fwd_trimmed))
sample_names <- gsub("[._-]$", "", sample_names)

cat("Samples:", length(sample_names), "\n")
```

---

# 5. Quality Filtering and Truncation

```{r}
#| label: filter-paths
fwd_filt <- file.path(dada2_dir, "filtered", paste0(sample_names, "_F_filt.fastq.gz"))
rev_filt <- file.path(dada2_dir, "filtered", paste0(sample_names, "_R_filt.fastq.gz"))
names(fwd_filt) <- sample_names
names(rev_filt) <- sample_names
```

```{r}
#| label: filter-and-trim
out <- filterAndTrim(
  fwd         = fwd_trimmed,
  filt        = fwd_filt,
  rev         = rev_trimmed,
  filt.rev    = rev_filt,
  truncLen    = truncLen,
  maxEE       = maxEE,
  truncQ      = truncQ,
  rm.phix     = rm_phiX,
  minLen      = min_length,
  compress    = TRUE,
  multithread = n_cores
)

rownames(out) <- sample_names
knitr::kable(out, caption = "Reads in/out after filterAndTrim")
```

```{r}
#| label: filter-check
# Remove samples with zero reads after filtering
kept <- out[, "reads.out"] > 0
if (any(!kept)) {
  warning("Samples with zero reads after filtering: ",
          paste(rownames(out)[!kept], collapse = ", "))
  fwd_filt     <- fwd_filt[kept]
  rev_filt     <- rev_filt[kept]
  sample_names <- sample_names[kept]
  out          <- out[kept, ]
}
```

---

# 6. Error Model Learning

## 6a. Continuous Q-scores (MiSeq / HiSeq)

Standard DADA2 parametric error model. Learn errors separately for forward
and reverse reads.

```{r}
#| label: error-learning-continuous
#| eval: !expr binned_quality
message("Learning error rates (continuous Q-scores)...")

err_fwd <- learnErrors(fwd_filt,
                        multithread = n_cores,
                        randomize   = TRUE)
err_rev <- learnErrors(rev_filt,
                        multithread = n_cores,
                        randomize   = TRUE)
```

## 6b. Binned Q-scores (NovaSeq / NextSeq)

NovaSeq and NextSeq use 2-channel chemistry that collapses quality scores into
a small number of bins (typically 4 bins: 2, 11, 25, 37). This violates
assumptions of the standard DADA2 error model.

**Strategy:** Use `LOESS` smoothing (`loessErrfun`) to flexibly model the
observed error rates instead of fitting a parametric function.

Reference: <https://github.com/ErnakovichLab/dada2_ernakovichlab>

```{r}
#| label: error-learning-binned
#| eval: !expr !binned_quality
message("Learning error rates (binned Q-scores – NovaSeq/NextSeq)...")

# For binned platforms, selfConsist iterates the error model to convergence;
# combine with loessErrfun for non-parametric error curve fitting.
err_fwd <- learnErrors(
  fwd_filt,
  errorEstimationFunction = loessErrfun,
  multithread  = n_cores,
  randomize    = TRUE,
  MAX_CONSIST  = 20,   # Extra iterations for convergence
  OMEGA_C      = 0     # Disable low-count smoothing (helps with binned data)
)

err_rev <- learnErrors(
  rev_filt,
  errorEstimationFunction = loessErrfun,
  multithread  = n_cores,
  randomize    = TRUE,
  MAX_CONSIST  = 20,
  OMEGA_C      = 0
)
```

## 6c. Error Model Diagnostics

```{r}
#| label: error-plots
#| fig-width: 10
#| fig-height: 5
p_err_fwd <- plotErrors(err_fwd, nominalQ = TRUE) +
  ggtitle("Forward read error model") +
  theme_bw()
p_err_rev <- plotErrors(err_rev, nominalQ = TRUE) +
  ggtitle("Reverse read error model") +
  theme_bw()

p_err_fwd / p_err_rev
```

::: callout-note
Points should roughly follow the lines (estimated error rates). Systematic
deviations may indicate a poor model fit, especially on binned-quality data.
:::

---

# 7. Sample Inference (Denoising)

DADA2 resolves true biological sequences from sequencing errors.

**Pooling options:**

- `"independent"` – fastest; each sample analysed alone
- `"pseudo"` – two-pass approach; improves sensitivity for rare variants
- `TRUE` – full pooling; maximum sensitivity, but memory-intensive

```{r}
#| label: denoise
message("Running DADA2 sample inference (pool = '", pool_method, "')...")

dada_fwd <- dada(fwd_filt, err = err_fwd,
                  multithread = n_cores,
                  pool        = pool_method)
dada_rev <- dada(rev_filt, err = err_rev,
                  multithread = n_cores,
                  pool        = pool_method)

# Summary of denoised sequences for first sample
dada_fwd[[1]]
```

---

# 8. Merge Paired Reads

```{r}
#| label: merge
message("Merging paired-end reads...")

mergers <- mergePairs(
  dadaF       = dada_fwd,
  derepF      = fwd_filt,
  dadaR       = dada_rev,
  derepR      = rev_filt,
  minOverlap  = min_overlap,
  maxMismatch = max_mismatch,
  verbose     = TRUE
)

# Fraction of reads successfully merged (first sample)
cat(sprintf("Sample 1 – Accepted mergers: %d / %d (%.1f%%)\n",
            sum(mergers[[1]]$accept),
            nrow(mergers[[1]]),
            100 * mean(mergers[[1]]$accept)))
```

---

# 9. ASV Table Construction

```{r}
#| label: make-seqtab
seqtab <- makeSequenceTable(mergers)
dim(seqtab)  # samples × ASVs
```

```{r}
#| label: seqlen-distribution
#| fig-width: 7
#| fig-height: 4
seq_lengths <- nchar(getSequences(seqtab))
seq_len_df  <- as.data.frame(table(seq_lengths))
colnames(seq_len_df) <- c("Length", "Count")
seq_len_df$Length <- as.integer(as.character(seq_len_df$Length))

ggplot(seq_len_df, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +
  geom_vline(xintercept = c(250, 255), linetype = "dashed", colour = "red") +
  labs(title = "ASV length distribution",
       subtitle = "Dashed lines indicate expected amplicon range",
       x = "Sequence length (bp)", y = "Number of ASVs") +
  theme_bw()
```

```{r}
#| label: trim-off-target
# Remove off-target length sequences (optional but recommended)
# Adjust the range to match your expected amplicon length
expected_range <- c(245, 260)  # bp – edit for your target region

seqtab_len <- seqtab[, nchar(colnames(seqtab)) %in% seq(expected_range[1], expected_range[2])]
cat(sprintf("ASVs before length filter: %d\n", ncol(seqtab)))
cat(sprintf("ASVs after  length filter: %d\n", ncol(seqtab_len)))
```

---

# 10. Chimera Removal

```{r}
#| label: remove-chimeras
message("Removing chimeras (method = '", chimera_method, "')...")

seqtab_nochim <- removeBimeraDenovo(
  seqtab_len,
  method      = chimera_method,
  multithread = n_cores,
  verbose     = TRUE
)

cat(sprintf("ASVs before chimera removal: %d\n", ncol(seqtab_len)))
cat(sprintf("ASVs after  chimera removal: %d\n", ncol(seqtab_nochim)))
cat(sprintf("Reads retained: %.1f%%\n", 100 * sum(seqtab_nochim) / sum(seqtab_len)))
```

---

# 11. Read Tracking

```{r}
#| label: read-tracking
getN <- function(x) sum(getUniques(x))

track <- cbind(
  out,
  sapply(dada_fwd, getN),
  sapply(dada_rev, getN),
  sapply(mergers,  getN),
  rowSums(seqtab_nochim)
)

colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample_names

knitr::kable(track, caption = "Read tracking through DADA2 pipeline")
```

```{r}
#| label: tracking-plot
#| fig-width: 9
#| fig-height: 5
track_df <- as.data.frame(track)
track_df$Sample <- rownames(track_df)

track_long <- tidyr::pivot_longer(
  track_df,
  cols      = -Sample,
  names_to  = "Step",
  values_to = "Reads"
)
track_long$Step <- factor(track_long$Step,
                            levels = c("input","filtered","denoisedF",
                                       "denoisedR","merged","nonchim"))

ggplot(track_long, aes(x = Step, y = Reads, group = Sample, colour = Sample)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Read retention through DADA2 pipeline",
       x = "Pipeline step", y = "Read count") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: low-retention-check
retention_pct <- 100 * track[, "nonchim"] / track[, "input"]
low_ret <- names(retention_pct[retention_pct < 50])
if (length(low_ret) > 0) {
  warning("Samples with < 50% read retention:\n  ",
          paste(low_ret, collapse = "\n  "))
} else {
  message("All samples retained >= 50% of reads. Good.")
}
```

---

# 12. Taxonomy Assignment

Taxonomy is assigned using DADA2's `assignTaxonomy()` function with the SILVA
(or GTDB) reference database, followed by `addSpecies()` for species-level
exact matching.

```{r}
#| label: assign-taxonomy
if (!file.exists(taxonomy_db)) {
  warning("Taxonomy database not found: ", taxonomy_db,
          "\nSkipping taxonomy assignment. Update `taxonomy_db` in User Parameters.")
  taxa <- NULL
} else {
  message("Assigning taxonomy (this may take several minutes)...")

  taxa <- assignTaxonomy(
    seqtab_nochim,
    refFasta    = taxonomy_db,
    multithread = n_cores,
    minBoot     = 80,    # Bootstrap confidence threshold
    verbose     = TRUE
  )

  if (file.exists(taxonomy_species)) {
    taxa <- addSpecies(taxa, taxonomy_species, verbose = TRUE)
  }

  cat("Taxonomy summary:\n")
  print(table(!is.na(taxa[, "Phylum"])))
}
```

```{r}
#| label: taxa-preview
if (!is.null(taxa)) {
  knitr::kable(
    head(taxa[, c("Kingdom","Phylum","Class","Order","Family","Genus")], 10),
    caption = "First 10 ASVs taxonomy"
  )
}
```

---

# 13. Save Outputs

```{r}
#| label: save-outputs
saveRDS(seqtab_nochim, file.path(dada2_dir, "seqtab_nochim.rds"))
if (!is.null(taxa)) saveRDS(taxa, file.path(dada2_dir, "taxa.rds"))
write.csv(track, file.path(dada2_dir, "track_reads.csv"))

# Save ASV table as a plain CSV (rows = samples, columns = ASV hashes)
asv_tab <- as.data.frame(t(seqtab_nochim))
rownames(asv_tab) <- paste0("ASV", seq_len(nrow(asv_tab)))
write.csv(asv_tab, file.path(dada2_dir, "asv_table.csv"))

# Save ASV sequences as FASTA
asv_seqs <- colnames(seqtab_nochim)
asv_ids   <- paste0("ASV", seq_along(asv_seqs))
asv_fasta <- file.path(dada2_dir, "asvs.fasta")
writeLines(paste0(">", asv_ids, "\n", asv_seqs), asv_fasta)

message("Outputs saved to: ", dada2_dir)
message("  seqtab_nochim.rds  – sequence table (samples × ASVs)")
message("  taxa.rds           – taxonomy matrix")
message("  track_reads.csv    – read tracking")
message("  asv_table.csv      – ASV counts (transposed)")
message("  asvs.fasta         – ASV sequences")
```

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
